Grow with the flow: a latitudinal cline in growth rate is associated with more variable precipitation in \textit{Mimulus cardinals}

### 2016-10-13 ms/ms.tex
ANOVA tables and more

% TO DO:
% 1. Is water regime on figure 1 correct?
% 2. Finalize Table 2
% 3. Do anything with biomass data?
% 4. Citations for latitudinal clines
	% cases where clinal variation has 'obvious' selective source:
	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
	% Caicedo et al 2004?
	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
% 5. Citations for GxE as signature of local adaptation
% 6. Fix figure 2
% 7. Supplemental Figure and Table captions
% 8. Supplemental figure showing agreement between assigned to climatic variables from point estimated and spatially averaged analyses

@@

% Local adaptation and clines
Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}. Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}. Phenotypic differences between populations along the cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}. Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons. In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment. We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change. A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures. Here, we examine physiological trait variation and possible selective agents in the perennail herb \textit{Mimulus cardinalis} Douglas ex Benth. (Phrymaceae), a model system for local adaptation studies.

% Intrinsic versus plastic variation
There are two basic approaches one can use to identify candidate traits underlying local adaptation in a common garden. First, if genetically-based trait differences between populations vary clinally with environmental differences, this may point to traits important for local adaptation. Second, genotype by environment interactions could indicate that variation in plasticity mediates local adaptation. We distinguish between these signatures of local adaptation by referring to `intrinsic' and `plastic' trait variation, respectively. There are classic cases adaptation involving both intrinsic and plastic trait variation. For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype by environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}.

% Selective agents
Either instrinsic and/or plastic variation should vary clinally along environmental gradients. Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely related to climatic variation. Despite the fact that latitudinal clines in particular have been studied for a long time, the exact climatic factors (e.g. minimum or maximum temperature, growing season length, strength of biotic interactions) is rarely known. Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g. coastal vs. continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}. In plants especially, we know little about the prevelance and adaptive significance of clinal variation in fundamental physiological traits like photosynthesis and the impact on plant performance.

@@

% Motivate primary questions using Mimulus
In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}. We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}. Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}. We measured genetic and genotype by environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude. We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity. Interannual variation in precipitaiton and/or precipitation seasonality is associated this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M. cardinalis}. We place these findings in the context of life history theory and consider future directions in the Discussion.

@@

We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements. We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.  We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing). Both leaf expansion and shoot elongation were modeled as a second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of 329 plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf \cdm{this is what I did...not sure exactly which node, as I didn't record that when I did measurements} acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), temperature, and moderate relative humidity. During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions.

\paragraph{Mortality}
We assayed mortality during twice-weekly growth measurements. We could not get GLMM with Family effects to converge, so we used GLM with a quasibinomial error structure and assessed signifiance using Type 2 Analysis of Deviance with the R package \pkg{car}.

@@

We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results). Since latitude \textit{per se} cannot be a selective agent, we considered three possible explanations. First, latitude may be strongly correlated, in this species, with one or two climatic variables, such as temperature, precipitation, or growing degree-days. Second, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latiduinal cline. Third, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration. For example, average temperature of \textit{M. cardinalis} populations at a given latitude varies widely, but in aggregate, a Southern metapopulation experiences warmer climate than a Northern one. Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g. high elevation) site.

To evaluate support for these hypotheses, we used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We looked for overlap between climatic variables that best predict latitude of \textit{M. cadinalis} occurence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations. For brevity, we refer to these as Climate-Latitude and Climate-Trait variables. We selected Climate-Latitude and Climate-Trait variables independently using Random Forest (VSURF algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}). From VSURF models, we kept only variables selected for prediction, the most stringent criterion.

The first hypothesis predicts that there should be one or two Climate-Latitude and Climate-Trait associations that are strongly correlated, whereas the second hypothesis predicts that several climatic variables should be weakly correlated. To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-Trait variables as above, except that we used spatially averaged climate variables. We sampled climate at 1000 random points (at 90-m resolition) within a 100-km buffer around focal populations. We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}. Since \textit{M. cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}. Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}. For clarity, we distinguish between analyses where climate is inferred from a single point (`point estimated Climate-Trait') versus averaged across a 100-km buffer (`spatially averaged Climate-Trait').

For these analyses, we compiled a representative set of 178 recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included 24 climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}). The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range. For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation. Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.

@@

Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M. cardinalis}, as evidenced by large and highly significant population effects (Table~\ref{table:Table_AnovaSummary}). A single principal component captured 74.2 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}). As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted greatly with little variation in plasticity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of Population, Treatment, Population x Treatment interactions for varies traits
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[ht]
   \centering
   \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects. We used different statistical modeling for the diverse traits assayed -- glm: generalized linear model using R \citep{R}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}. Note that temperature and water treatments were imposed after germination, hence are not application to this trait. Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}
   \resizebox{6.5in}{!}{
   \begin{tabular}{@{} llllllll @{}}
      \toprule
    {\raggedleft Trait}                          & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
    Statistical model              & survreg     & lmer           & lmer            & lmer           & lmer          & glm \\
      \midrule
  Population                       &
    ***  &
    *** &
    *** &
    *** & 
    *** & 
    *** & \\
  Temperature                      &
    NA                                                          &
    ***    &
    *** &
    *** & 
     &
    *** & \\
  Water                            &
    NA                                                          &
    ***   &
    ***   &
    *** & 
     &     
    *** & \\
  Pop $\times$ Temp                &
    NA                                                          &
     &
     &
    * & 
     &     
    * & \\
  Pop $\times$ Water               &
     NA                                                          &
    * &
     &
     & 
     &     
     & \\
  Temp $\times$ Water              &
    NA                                                          &
     &
    *** &
     & 
     &     
    *** & \\
  Pop $\times$ Temp $\times$ Water &
    NA                                                          &
     &
     &
     & 
     &     
     & \\
      \bottomrule
   \end{tabular} }
   \label{table:Table_AnovaSummary}
\end{table}

@@

\begin{figure}[h!]
	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude. Each point is a population, its latitude of origin a position along the slow to fast growth axis, defined as Principal Component 1 of five traits (see Methods). The line and 95\% confidence intervals were estimated using standard linear regression.}
	\label{fig:Fig_PC1vLat}
\end{figure}

\subsection*{Remarkably little evidence for variation in plasticity}

Genotype $\times$ environment interactions are also a common signature of local adaptation. For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions. We found remarkably little evidence for this pattern in \textit{M. cardinalis}. There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing. Otherwise, populations responded remarkably similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}). Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects. However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.

>>>>

%lost on page 11 - extracted and used climate variables
%Maybe reordering stuff on paragraph
%cartoon of whole procedure for climate variables
%figure showing correlation b/t bio12 and 
%state that VSURF importance is different scales in different models
%moved up three hypotheses part to introduction
%play up lack of GxE and ???
%** correct # of Pop x Treatment

%* delve into interactions - do they change cline?

%* show figures of reaction norms
%* split plot, account for error variation in temp???
%* show PAR similarity
%* some sites they are functionally annual

%* hung up on specialization - clarify? where is ecological oppportunity, tradeoffs, specialization
%	- maybe temporal source-sink makes more sense

%* explain rhizomes

%* put in biomass allocation cline
% TO DO:
% 1. Is water regime on figure 1 correct? (data is in office)

% 3. Do anything with biomass data?
% 4. Citations for latitudinal clines
	% cases where clinal variation has 'obvious' selective source:
	% Antonovics and Bradshaw discuss steepness of clines and strength of selection over 10's of meters off mines in Anthoxanthum
	% Caicedo et al 2004?
	% clines in phenotypic traits reviewed by Hedrick 2006, Schemske and Bier 2007
% 5. Citations for GxE as signature of local adaptation
% 7. Supplemental Figure and Table captions
% 8. Supplemental figure showing comparison between assigned to climatic variables from point estimated and spatially averaged analyses
% 9. new spatially averaged analysis
% 12. For Figure_PC1vLat: plot axes could be range maps with areas highlighted (x-axis) and short to tall plants (y-axis)

@@

% Local adaptation and clines
Local adaptation within species is ubiquitous; populations generally have higher fitness in their native environment, but perform poorly outside it \citep{Schluter_2000, Hereford_2009}. Local adaptation also frequently leads to clines in both phenotypes and allele frequencies when selection varies over environmental gradients \citep{Huxley_1938, Endler_1977}. Phenotypic differences between populations along a cline most often have a genetic basis and can be studied in a common garden \citep{Turesson_1922, Clausen_etal_1940, Hiesey_etal_1942}. Despite a long history of studying local adaptation and clines, it remains to challegenging to identify exactly which traits are under selection and which differ for nonadaptive reasons. In particular, the role that physiological differences play in local adaptation is poorly understood, despite the fact that physiology is frequently assumed to explain adaptation to the abiotic environment. We need to understand physiological adaptations within species as a baseline for anticipating how organisms will respond to climate change. A related problem is identifying which features of the environment, abiotic factors like soil water availability or biotic interactions, cause spatially varying selective pressures. Here, we examine physiological trait variation and possible selective agents in the perennial herb \textit{Mimulus cardinalis} Douglas ex Benth. (Phrymaceae), a model system for local adaptation studies.

% Intrinsic versus plastic variation 
When populations are locally adapted, reaction norms for fitness will cross, such that local genotypes have higher fitness than foreign genotypes and rank orders change across environments \citep{Kawecki_Ebert_2004} \ala{added citation - might give error}. The traits that underlie local adaptation, however, need not mirror this pattern. Populations can have fixed genetic differences conferring trait values that are adaptive at home but neutral or maladaptive away. Alternatively, genotype-by-environment interactions could indicate that variation in plasticity mediates local adaptation. We distinguish between these patterns of adaptive trait differences by referring to `intrinsic' and `plastic' trait variation, respectively. There are classic cases of adaptation involving both intrinsic and plastic trait variation. For example, intrinsic differences in critical photoperiod [CITE] and developmental rate \citep{Stinchcombe_etal_2004} allow organisms to properly time their life history with the local environment. Conversely, sun and shade plants do not have intrinsically higher or lower rates of carbon assimilation, but rather, genotype-by-environment interactions cause sun plants to assimilate more under high light and shade plants under low light \citep{Givnish_1988}. In plants especially, we know little about the prevalance and adaptive significance of variation in fundamental physiological traits like photosynthesis and their impact on plant performance.

% Selective agents (trait-enviro assoc)
A basic approach one can use to identify candidate traits underlying local adaptation is to find associations between traits and environments. Either instrinsic and/or plastic variation should vary clinally along environmental gradients. Indeed, clines in ecologically important traits are widespread in nature \citep{Endler_1977} and often adaptive, but in most cases the selective agent is unknown. For example, in \textit{Drosophila} numerous latitudinal clines exist for traits like thermal tolerance \citep{Hoffmann_etal_2002}, body size (\cite{Coyne_Beecham_1987} and references therein), and life history \citep{Schmidt_etal_2005}. Some \textit{Drosophila} clines have evolved multiple times (\cite{Oakeshott_etal_1982, Huey_etal_2000}, see also \cite{Bradshaw_Holzapfel_2001}) or shifted in response to climate change \citep{Umina_etal_2005}, evincing climatic adaptation. Similarly, plant species exhibit latitudinal clines in traits like flowering time \citep{Stinchcombe_etal_2004}, cyanogenesis \citep{Kooyers_Olsen_2012}, and leaf morphology \citep{Hopkins_etal_2008} that likely relate to climatic variation.

Despite the fact that latitudinal clines in particular have been studied for a long time, latitude \textit{per se} cannot be a selective agent. Latitude may be strongly correlated with one or two key climatic variables, such as temperature, precipitation, or growing degree-days, such that latitude is an effective proxy for the underlying climatic driver but nonetheless  key climatic variable(s) with stronger relationships to traits exist. Alternatively, latitude may be more strongly related to traits than any single climatic variable for two possible reasons. First, latitude may be correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline. Alternatively, gene flow among neighboring populations could smooth out local climatic effects, since alleles will experience selection across populations linked by migration. For example, in mountainous regions average temperature at a given latitude varies widely, but in aggregate, a Southern set of populations will experience warmer climate than a Northern one. Thus, any particular Southern population would be warm-adapted, even if it was located in cooler (e.g. high elevation) site. Because many climatic factors vary latitudinally, and which climatic factors vary latitudinally changes over the earth's surface (e.g. coastal vs. continental), dissecting the evolution of latitudinal clines across many species will help biologists identify generalities, such as whether thermal tolerance maxima or seasonal timing is more important \citep{Bradshaw_Holzapfel_2008}, or whether local or regionally-averaged variables are more predictive of trait change. \ala{added this last clause to try to encompass gene flow idea, but now maybe the sentence is too clunky}

@@

% Motivate primary questions using Mimulus
% might add that Anacker and Strauss 2014 showed ecological shifts important in Mim speciation and we would like to know driving forces behind this???
In this study, we address these gaps by asking whether intrinsic or plastic physiological trait variation corresponds with latitude and what climatic factor(s) could plausibly be repsonsible for latitudinal clines in a focal species, \textit{Mimulus cardinalis}. We chose this species because linking physiological traits to potentially complex patterns of local adaptation requires integrating multiple lines of evidence from comparative, experimental, and genomic studies under both lab and field conditions. Many classic and contemporary studies of local adaptation use species from genus \textit{Mimulus} because of its natural history, easy propogation, and genetic/genomic resources \citep{Clausen_etal_1940, Hiesey_etal_1971, Bradshaw_Schemske_2003, Wu_etal_2008, Lowry_Willis_2010, Wright_etal_2013}. Yet, there is a conspicuous deficieny of links between local adaptation and physiological mechanisms \citep{Angert_2006, Angert_etal_2008, Wu_etal_2010}. We measured genetic and genotype-by-environment variation in response to temperature and drought among 16 populations distributed over 10.7\textdegree of latitude. We found a latitudinal cline of intrinsic differences in photosynthesis and growth, but no evidence for variation in plasticity. Interannual variation in precipitation and/or precipitation seasonality is associated with this axis of variation, suggesting that climatic variance rather than mean may be an important driver of local adaptation in \textit{M. cardinalis}. We place these findings in the context of life history theory and consider future directions in the Discussion. \ala{wrap spatial buffer results into summary?}

@@

We measured growth rate in two phases: leaf expansion as a rosette and shoot elongation after bolting. We censused leaf length twice per week from 12 May -- 12 June (28--59 days after sowing), resulting in 10 measurements. We ceased measuring leaf length once it appeared to asymptote and growth shifted to shoot elongation.  We also censused plant height on 7 days (twice per week) between 29 May and 20 June (45 to 67 days after sowing). Both leaf expansion and shoot elongation were modeled as second-order polynomials of time with individual coefficients (separate for leaf and shoot growth) using empirical Bayes' estimates from linear mixed-effects models fit using the R package \pkg{lme4} version 1.1-7 \citep{Bates_etal_2014}.

\paragraph{Photosynthesis}
During the week of 10 to 16 June (57 to 63 days after sowing), we measured daytime photosynthetic rate and stomatal conductance on a subset of 329 plants evenly spread between treatments and families within populations. The youngest, fully-expanded leaf acclimated for 3 minutes to reach steady state in a 6 cm$^2$ chamber of a LI-COR 6400XT Portable Photosynthesis System (LI-COR Biosciences, Lincoln, Nebraska). All measurements were made at ambient light (400 $\mu$mol m$^{-2}$ s$^{-1}$), atmospheric CO$_2$ (400 ppm), temperature, and moderate relative humidity. During this period, we suspended normal day-to-day temperature fluctuations and set daytime temperatures to its average for that period (Cool: 26.5$\degree$; Hot: 36.1$\degree$ so that all plants within a temperature level were measured under the same conditions. As expected, stomatal conductance was strongly correlated with photosynthetic rate (results not shown), but we found that treatment and population effects were more apparent when stomatal conductance was factored out. Therefore, we calculated the `intrinsic' photosynthetic rate for each individual as its residual deviance from the regression between stomoatal conductance and photosynthetic rate across all plants. We also removed one outlier.

\paragraph{Mortality}
We assayed mortality during twice-weekly growth measurements. We analyzed the probability of surviving until the end of the experiment as a function of population, treatment, and their interactions using a Generalized Linear Mixed Model (GLMM) assuming binomially distributed errors. We included Family and Block as random effects. We assessed significance of fixed effects using Type 2 Analysis of Deviance with Wald $\chi ^2$ tests in the R package \pkg{car} \citep{Fox_Weisberg_2011}.

@@

We found that a population's position along PC1 correlated strongly with the latitude or origin (see Results) and next used Random Forest regression \citep{Liaw_Wiener_2002} to identify putative climatic factors underlying trait-latitude associations in \textit{M. cardinalis}. We looked for overlap between climatic variables that best predict latitude of \textit{M. cardinalis} occurrence records and a separate analysis of climatic variables that best predict trait variation across our 16 focal populations. For brevity, we refer to these as `Climate-Latitude' and `Climate-TraitPC1' variables. We selected Climate-Latitude and Climate-TraitPC1 variables independently with Variable Selection Using Random Forest (VSURF) algorithm in the R \pkg{VSURF} version 0.8.2 \citep{Genuer_etal_2014}. From VSURF models, we kept only variables selected for prediction, the most stringent criterion.

If latitude is strongly correlated with one or two climatic variables that are responsible for the cline, then there should be one or two climatic variables that are identified in both the Climate-Latitude and Climate-TraitPC1 analyses \ala{and the correlation magnitudes should be high? maybe this language is not technically correct (variable importance more accurate?) but the point is that this prediction needs something about count and the strength}. If latitude is correlated with several climatic agents of selection that are individually weak, but add up to a strong latitudinal cline, then several climatic variables should be weakly correlated \ala{fix this to mirror prior}.

To test the third hypothesis about gene flow smoothing out local climatic variation, we repeated the same procedure to identify Climate-TraitPC1 variables as above, except that we used spatially averaged climate variables. \ala{what's the actual prediction? that spatially averaged Climate-TraitPC1 will have higher correlation than point estimated Climate-Trait?} We sampled climate at 1000 random points (at 90-m resolition) within a 62-km buffer around focal populations. We chose this buffer size because neutral genetic differentiation increases slowly with geographic distance, indicating significant gene flow between nearby populations \citep{Paul_etal_2016}. Significant spatial autocorrelation persisted for approximately 62 km. Since \textit{M. cardinalis} is found exclusively in riparian areas, we only selected points along streams using the National Hydrogeoraphy Dataset \citep{NHD}. Climatic means and CVs were weighted by their climatic suitability as determined using a multimodel ensemble average of ecological niche models \citep{Angert_ENM}. We distinguish between analyses where climate is inferred from a single point (`point estimated Climate-TraitPC1') versus averaged across a 62-km buffer (`spatially averaged Climate-TraitPC1').

For these analyses, we compiled a representative set of 178 recent (since 2000) known \textit{M. cardinalis} occurences. These occurences were thinned by 50\% to correct for uneven sampling from a comprehensive set of herbarium records and an exhaustive field survey in 2010-11 \citep{Angert_ENM}. For each occurence, we used a 90-m digital elevation model from HydroSHEDS \citep{Lehner_etal_2006} to extract elevation. Monthly interpolated climate layers were calculated using ClimateWNA version 5.30 \citep{Wang_etal_2012}, which accurately downscales climate data specifically for the rugged topography of western North America. For each occurence, we calculated bioclimatic variables using the biovars function in the R package \pkg{dismo} \citep{Hijmans_etal_2014}. In total, we included 24 climate variables, 9 from ClimateWNA and 15 bioclimatic variables (Table~\ref{table:TableS_ClimVars}). The bioclimatic variables included all permutations of two climatic factors, temperature and precipitation, and six temporal scales (annual average, coldest quarter, warmest quarter, wettest quarter, driest quarter, or seasonality) as well as mean diurnal range, isothermality, annual temperature range. For each variable, we calculated both a 30-year normal by averging annual values between 1981 and 2010 and 30-year coefficient of variation, a standardized metric of interannual climatic variation. Temperatures were converted to Kelvin to be on a ratio scale appropriate for calculating the coefficient of variation.

@@

Using a common garden, we identified strong genetically-based trait differences in time to germination, growth, and photosynthetic rate among populations of \textit{M. cardinalis}, as evidenced by large and significant population effects (Table~\ref{table:Table_AnovaSummary}). A single principal component captured 74.4 \% of the trait variation among populations, defining an axis of variation from fast to slow growth (Fig~\ref{fig:Fig_PC1vLat}). As we explain below, intrinsic differences between populations in terms of plant function (photosynthesis) and performance (growth) contrasted with the low amount of variation in plasticity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of Population, Treatment, Population x Treatment interactions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[ht]
   \centering
   \topcaption[ANOVA summary]{Summary of Population, Treatment, and Population $\times$ Treatment effects. We used different statistical modeling for the diverse traits assayed -- glmer: generalized linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; lmer: linear mixed model using the R package \pkg{lme4} \citep{Bates_etal_2014}; survreg: survival regression using the R package \pkg{survival} \citep{Therneau_2015}. Note that temperature and water treatments were imposed after germination, hence are not applicable to this trait. Complete analysis of variance/deviance tables for each trait are available in the Supporting Information. Key to statistical significance: *$P < 0.05$; ** $P < 0.01$; *** $P < 0.001$}
   \resizebox{6.5in}{!}{
   \begin{tabular}{@{} llllllll @{}}
      \toprule
    {\raggedleft Trait}                          & Germination & Leaf expansion & Shoot elongation & Photosynthesis & Intrin. Photo & Mortality \\
    Statistical model              & survreg     & lmer           & lmer            & lmer           & lmer          & glmer \\
      \midrule
  Population                       &
    ***  &
    *** &
    *** &
    *** & 
    *** & 
     & \\
  Temperature                      &
    NA                                                          &
    ***    &
    *** &
    ** & 
     &
    *** & \\
  Water                            &
    NA                                                          &
    *   &
       &
     & 
     &     
    *** & \\
  Pop $\times$ Temp                &
    NA                                                          &
     &
     &
    * & 
     &     
     & \\
  Pop $\times$ Water               &
     NA                                                          &
    * &
     &
     & 
     &     
     & \\
  Temp $\times$ Water              &
    NA                                                          &
     &
     &
     & 
     &     
    *** & \\
  Pop $\times$ Temp $\times$ Water &
    NA                                                          &
     &
     &
     & 
     &     
     & \\
      \bottomrule
   \end{tabular} }
   \label{table:Table_AnovaSummary}
\end{table}

@@

\begin{figure}[h!]
	\centerline{\includegraphics[width=1\textwidth]{Figures/Figure_PC1vLat.pdf}}
	\usefont{T1}{phv}{m}{n}
	\fontsize{10}{12}
	\selectfont
	\caption[Southern populations grow faster]{Trait variation, from fast to slow growth, is closely associated with latitude. Each point is a population's latitude of origin (x-axis) and position along the slow to fast growth axis (y-axis), defined as Principal Component 1 of five traits (see Methods). The line and 95\% confidence intervals were estimated using linear regression.}
	\label{fig:Fig_PC1vLat}
\end{figure}

\subsection*{Little evidence for variation in plasticity}

Genotype $\times$ environment (G$\times$E) interactions are also a common signature of local adaptation. For example, populations from more stressful environments may tradeoff reduced growth rate under benign conditions for the ability to tolerate stress by maintaining positive growth rates and/or surviving through adverse conditions. We found little evidence G$\times$E in \textit{M. cardinalis}. \cdm{Is this result still correct after revising analysis?} There was only one statistically significant Population $\times$ Treatment interaction (Table~\ref{table:Table_AnovaSummary}), but this idiosyncratic result was weak and would not have survived correction for multiple testing. Otherwise, populations responded similarly to treatments: faster growth in the hot treatment, slower growth in the dry treatment, and high mortality in the hot, dry treatment (Table~\ref{table:Table_AnovaSummary}). Note that interactions were calculated after factoring out intrinsic trait differences, necessarily reducing statistical power to detect significant interactions relative to main effects. However, the fact that the Population and Treatment effects were highly significant ($P \ll 0.001$ in most cases) suggests that statistical power alone cannot explain why we failed to detect Population $\times$ Treatment interactions.
